Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-11 image_analysis_paper_development2/SI_text.tex
updated figures to make compiled panels

\begin{figure}
\includegraphics[width=8cm]{drosophila_live_eval_spectrum}

\includegraphics[width=2.5cm]{drosophila_live_evec_corr1}
\includegraphics[width=2.5cm]{drosophila_live_evec_corr2}
\includegraphics[width=2.5cm]{drosophila_live_evec_corr3}
\caption{Eigenvalue spectra for the {\em Drosophila} live imaging data set presented in \fig~3. Note that there is a gap after the fourth eigenvalue product. Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate. Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics[width=8cm]{zebrafish_eval_spectrum}

\includegraphics[width=2.5cm]{zebrafish_evec_corr1}
\includegraphics[width=2.5cm]{zebrafish_evec_corr2}
\includegraphics[width=2.5cm]{zebrafish_evec_corr3}
\caption{Eigenvalue spectra for the zebrafish data set presented in \fig~4. Note that there is a gap after the fourth eigenvalue product. Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate. Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  It is not immediately obvious that coordinate 3 is a harmonic of coordinate 1; the distortion in the plot is due to density effects in the data (the developmental changes are slower towards the beginning of the trajectory, and so there is a higher density of images in this portion of the one-dimensional curve), We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics[width=8cm]{drosophila_fixed_eval_spectrum}
\caption{Eigenvalue spectra for the fixed {\em Drosophila} images presented in \fig~5. Note that there is a gap after the first eigenvalue product. We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics[width=8cm]{wing_disc_eval_spectrum}
\caption{Eigenvalue spectra for the wing disc data set presented in \fig~6. Note that there is a gap after the first eigenvalue product. We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\textsf{
\begin{tabular}{| r | c | c |}
\hline
Data set & \parbox[c]{4cm}{ \centering VDM Rank Correlation} & \parbox[c]{4cm}{ \centering PCA Rank Correlation} \\
\hline
{\em Drosophila} gastrulation (live) & 0.9989 & 0.8137 \\
Zebrafish epiboly & 0.9955 & 0.6351 \\
{\em Drosophila} gastrulation (fixed) & 0.9716 & 0.8658 \\
{\em Drosophila} wing discs & 0.9436 & 0.9381 \\
\hline
\end{tabular}
}
\caption{Comparison between rank correlation coefficients when ordering using the first vector diffusion maps (VDM) embedding coordinate, and rank correlation coefficients when ordering by the first principal component analysis (PCA) \citep{shlens2005tutorial} projection coefficient. For ordering using PCA, we computed the first principal component of the registered images, and then ordered the images by the projection coefficients onto this first mode. The PCA ordering is always less accurate than the vector diffusion maps ordering. In the zebrafish data set, the ordering is much less accurate using PCA, as the dynamics of the morphing and spreading of the cell mass are highly nonlinear. The PCA and VDM orderings are comparable for the wing disc data; this is to be expected, as the tissue simply grows in time.  }
\end{figure}

>>>>

\begin{figure}
\includegraphics{figS3}
\caption{Eigenvalue spectra for the {\em Drosophila} live imaging data set presented in \fig~3. Note that there is a gap after the fourth eigenvalue product. Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate. Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics{figS4}
\caption{Eigenvalue spectra for the zebrafish data set presented in \fig~4. Note that there is a gap after the fourth eigenvalue product. Below are the second, third, and fourth embedding coordinate plotted versus the first embedding coordinate. Note that coordinates 2--4 are higher harmonics (and thus simple functions) of coordinate 1, and are therefore not informative about structure in the data set.  It is not immediately obvious that coordinate 3 is a harmonic of coordinate 1; the distortion in the plot is due to density effects in the data (the developmental changes are slower towards the beginning of the trajectory, and so there is a higher density of images in this portion of the one-dimensional curve), We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics{figS5}
\caption{Eigenvalue spectra for the fixed {\em Drosophila} images presented in \fig~5. Note that there is a gap after the first eigenvalue product. We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{figure}
\includegraphics{figS6}
\caption{Eigenvalue spectra for the wing disc data set presented in \fig~6. Note that there is a gap after the first eigenvalue product. We can conclude that the data set is effectively one-dimensional and can be parameterized/ordered by coordinate 1.}
\end{figure}

\begin{table*}
\textsf{
\begin{tabular}{| r | c | c |}
\hline
Data set & \parbox[c]{4cm}{ \centering VDM Rank Correlation} & \parbox[c]{4cm}{ \centering PCA Rank Correlation} \\
\hline
{\em Drosophila} gastrulation (live) & 0.9989 & 0.8137 \\
Zebrafish epiboly & 0.9955 & 0.6351 \\
{\em Drosophila} gastrulation (fixed) & 0.9716 & 0.8658 \\
{\em Drosophila} wing discs & 0.9436 & 0.9381 \\
\hline
\end{tabular}
}
\caption{Comparison between rank correlation coefficients when ordering using the first vector diffusion maps (VDM) embedding coordinate, and rank correlation coefficients when ordering by the first principal component analysis (PCA) \citep{shlens2005tutorial} projection coefficient. For ordering using PCA, we computed the first principal component of the registered images, and then ordered the images by the projection coefficients onto this first mode. The PCA ordering is always less accurate than the vector diffusion maps ordering. In the zebrafish data set, the ordering is much less accurate using PCA, as the dynamics of the morphing and spreading of the cell mass are highly nonlinear. The PCA and VDM orderings are comparable for the wing disc data; this is to be expected, as the tissue simply grows in time.  }
\end{table*}

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
updated figures to make compiled panels

\begin{figure*}[t]
\raisebox{0.75cm}{\textsf{A}}
\includegraphics[width=17cm]{drosophila_live_imaging_scrambled}

\raisebox{0.75cm}{\textsf{B}}
\includegraphics[width=17cm]{drosophila_live_imaging_ordered}

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\
2 & 4.98$^{\circ}$ & 0.9994 \\
3 & 6.47$^{\circ}$ & 0.9795 \\
4 & 7.03$^{\circ}$ & 0.9953 \\
5 & 3.04$^{\circ}$ & 0.9959 \\
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation. Scale bar indicates 50$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A registered and ordered by vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

\begin{figure*}[t]
\begin{minipage}[c]{13.75cm}
\vspace{0cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=13cm]{zebrafish_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=13cm]{zebrafish_ordered}
\end{minipage}
%
\hfill
%
\begin{minipage}[c]{3.75cm}

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a movie of zebrafish epiboly. Scale bar indicates 200$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

@@

\begin{figure*}[t!]
\begin{minipage}[b]{12cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}

\raisebox{0.5cm}{\textsf{B}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_ordered}

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analysis of images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).  Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

@@

\begin{figure*}[t]
\raisebox{5.5cm}{\textsf{A}}
\includegraphics[width=6cm]{wing_disc_example}
%
\hfill
%
\raisebox{5.5cm}{\textsf{B}}
\includegraphics[width=10cm]{wing_disc_ordered}

\vspace{0.25cm}
\raisebox{2.25cm}{\textsf{C}}
\includegraphics[width=17.5cm]{wing_disc_average}

\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks. (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack. The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated. Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray). Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

@@

\begin{figure*}[t]
\raisebox{5cm}{\textsf{A}}
\includegraphics[width=5.5cm]{timing_studies_nimages}
%
\raisebox{5cm}{\textsf{B}}
\includegraphics[width=5.5cm]{timing_studies_npixels}
%
\raisebox{5cm}{\textsf{C}}
\includegraphics[width=5.5cm]{timing_studies_nrot}

\raisebox{2.5cm}{\textsf{D}}
{\renewcommand{\arraystretch}{1.5}
{ \textsf{ \small
\begin{tabular}{ | p{2cm} | c | c | c | c | c | c | }
\hline
Data Set &
Data Type &
\parbox[c]{2cm}{ \centering Number of Channels} &
\parbox[c]{2cm}{ \centering \vspace{.5\baselineskip} Number of Images \vspace{.5\baselineskip}} &
\parbox[c]{2cm}{ \centering Number of Pixels} &
\parbox[c]{2cm}{ \centering Angular Discretization} &
\parbox[c]{2cm}{ \centering CPU time}  \\
\hline
{\em Drosophila} gastrulation (live) & 2D & 1 & 40 & 100 $\times$ 100 & 10$^\circ$ & 3.2~sec \\
Zebrafish epiboly & 2D & 1 & 120 & 100 $\times$ 100 & 10$^\circ$ & 13~sec\\
{\em Drosophila} gastrulation (fixed) & 2D & 3 & 120 & 100 $\times$ 100 & 10$^\circ$ & 29~sec  \\
{\em Drosophila} wing discs & 3D & 3 & 46 & 100 $\times$ 100 $\times$ 21 & 10$^\circ$ & 12~sec  \\
 \hline
\end{tabular}}}}
\caption{Computational requirements for the presented methodology. (A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ angular discretization). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images. (B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in the number of pixels. (C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels). Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization. (D) The algorithm settings and computational requirements for the data sets analyzed. All times are reported for an Intel Core i7 2.93 GHz processor.}
\label{fig:cpu_time}
\end{figure*}

>>>>

\begin{figure*}[t]
\includegraphics{fig3}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation. Scale bar indicates 50$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A registered and ordered by vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

\begin{figure*}[t]
\includegraphics{fig4}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a movie of zebrafish epiboly. Scale bar indicates 200$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

@@

\begin{figure*}[t!]
\includegraphics{fig5}
%
\caption{Analysis of images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).  Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

@@

\begin{figure*}[t]
\includegraphics{fig6}
\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks. (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack. The anterior-posterior (A-P) and dorsal-ventral (D-V) axes are indicated. Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray). Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

@@

\begin{figure*}[t]
\includegraphics{fig7}
\caption{Computational requirements for the presented methodology. (A) CPU time as a function of the number of images in the data set (for $100 \times 100$ pixel images, and $10^\circ$ angular discretization). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.33})$ in number of images. (B) CPU time as a function of the number of pixels in the images (for $120$ images, and $10^\circ$ angular discretization). Empirically, the CPU time is $\sim \mathcal{O}(n^{1.83})$ in the number of pixels. (C) CPU time as a function of the number of rotations (for $120$ images of $100 \times 100$ pixels). Empirically, the CPU time is $\sim \mathcal{O}(n^{-0.77})$ in the angular discretization. (D) The algorithm settings and computational requirements for the data sets analyzed. All times are reported for an Intel Core i7 2.93 GHz processor.}
\label{fig:cpu_time}
\end{figure*}
